Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-19 co-regular-EPQ.tex
modify

\begin{restatable}{lemma}{EPQasMultiInMRpriforHistory}
\label{lemma:EPQ as multi in MRpri for history}
Given a data-differentiated execution $e$, $e \sqsubseteq \seqPQ$, if and only if, $\forall e' \in \textit{proj}(e)$, $\Gamma\mathsf{\text{-}Condition}(e) \Rightarrow \Gamma\mathsf{\text{-}Conc}(e,\alpha)$ holds for some $\alpha$
\end{restatable}

@@

Note that due to data-independence, register automata are used to detect some renamed violations, instead of store and checking directly.

Let us introduce the notion of co-regular, which reduce checking linearizable w.r.t $\Gamma\mathsf{\text{-}Conc}(e,\alpha)$ into checking set of witness automata.

%\vspace{-6pt}
\begin{definition}\label{def:co-regular of rules of extended priority queues}
Given $R \in \{ \textit{EPQ}_1^{>}, \textit{EPQ}_1^{=}, \textit{EPQ}_2^{>}, \textit{EPQ}_2^{=}, \textit{EPQ}_3 \}$. $R$ is co-regular, if there are a finite set $\textit{Auts}_{R}$ of witness automata such that, for each data-independence implementation $\mathcal{I}$, we have that

$$\textit{Auts}_{R} \cap \mathcal{I} \neq \emptyset \Leftrightarrow \exists e \in \mathcal{I}_{\neq},e' \in \textit{proj}(e), R \in last(e') \wedge e' \not\sqsubseteq \textit{MS}(R)$$

We say that $\textit{EPQ}$ is co-regular, if all five candidates of $R$ are co-regular.
\end{definition}

We simply the proof of co-regular of $\textit{EPQ}$ in two side. The first side uses the results in \cite{Bouajjani:2015} to ensure FIFO (first in first out) property of each single-priority projections. A execution where only items of one priority is putted is called a single-priority execution. It is obvious that each single-priority projection of sequences in $\textit{EPQ}$ satisfy FIFO property. Let $\textit{transToQueue}(e)$ be an execution generated from $e$ by transforming $\textit{put}$ and $\textit{rm}$ into $\textit{enq}$ and $\textit{deq}$, respectively, and then discarding priorities. In Appendix \ref{sec:appendix proof and definition in section definition of co-regular}, according to the result of queue in \cite{Bouajjani:2015}, we construct a set $\textit{Auts}_{\textit{sinPri}}$ of witness automata, and shows that they are enough to ensure that for each data-differentiated executions, each of its single-priority projection without $\textit{rm}(\textit{empty})$ to have ``FIFO'' property, as shown by the following lemma.

@@

According to Lemma \ref{lemma:automata for extended priority queue with single priority}, from now on, it is safe to assume that, for each data-differentiated execution without $\textit{rm}(\textit{empty})$, any of its single-priority projection has ``FIFO'' property. For example, $\textit{rm}(a)$ never happens before $\textit{put}(a,\_)$ for each $a$.

The second side is to ignore the incomparable priorities. Given a data-differentiated execution $e$, we say that $e$ is a $\textit{pri}$-execution, if $\textit{pri}$ is the maximal priority of $e$, and $\textit{pri}$ is larger than or equal to the priority of all other items of $e$. Given a data-differentiated execution $e$ and assume that $\textit{pri}$ is one of the maximal priority of $e$, let $\textit{pri-Exec}(e)$ be an execution obtained from $e$ by erasing all operations of items whose priority are incomparable with $\textit{pri}$. The following lemma states that when checking co-regular of $\textit{EPQ}$, it is safe to consider $\textit{pri}$-executions. We can see that $\textit{last}(e')$ contains at most one element for $\textit{pri}$-execution $e'$.

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Given a data-differentiated execution $e$ and $R = \textit{EPQ}_1^{>}$ (resp., $R = \textit{EPQ}_1^{=}$, $R=\textit{EPQ}_2^{>}$, or $R = \textit{EPQ}_2^{=}$). $e \sqsubseteq \textit{MS}(R)$ with witness $x$, if and only if $\textit{pri-Exec}(e) \sqsubseteq \textit{MS}(R)$ with witness $x$, where $\textit{pri}$ is the priority of $x$. %for each $\textit{pri}$ that is (1) maximal in $e$ and (2) items of $\textit{pri}$ are only one pair of matched $\textit{put}$ in $e$ (resp., more than one pair of matched $\textit{put}$, unmatched $\textit{put}$, or unmatched $\textit{put}$ and matched $\textit{put}$).
\end{restatable}

>>>>

\begin{restatable}{lemma}{EPQasMultiInMRpriforHistory}
\label{lemma:EPQ as multi in MRpri for history}
Given a data-differentiated execution $e$, $e \sqsubseteq \seqPQ$, if and only if, $\forall e' \in \textit{proj}(e)$, $\Gamma\mathsf{\text{-}Condition}(e) \Rightarrow \Gamma\mathsf{\text{-}Conc}(e,\alpha)$ holds for some $\alpha$ for each $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMaxPriority}$ $,\mathsf{MatchedMaxPriority}\}$. 
\end{restatable}

@@

Note that due to data-independence, register automata are used to detect some renamed violations, instead of store and checking directly. Let us introduce the notion of co-regular, which reduce checking linearizable w.r.t $\Gamma\mathsf{\text{-}Conc}(e,\alpha)$ into checking set of witness automata.

%\vspace{-6pt}
\begin{definition}\label{def:co-regular of rules of extended priority queues}
$\seqPQ$ is co-regular, if for each $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMaxPriority}$ $,\mathsf{MatchedMaxPriority}\}$, there is a finite set $\textit{Auts}_{\Gamma}$ of register automata, such that for each data-independence implementation $\mathcal{I}$, we have that:

$\textit{Auts}_{\Gamma} \cap \mathcal{I} \neq \emptyset$ if and only if $\exists e \in \mathcal{I}_{\neq},e' \in \textit{proj}(e)$, such that $\Gamma\mathsf{\text{-}Condition}(e')$ holds but $\Gamma\mathsf{\text{-}Conc}(e',\alpha)$ does not hold for some $\alpha$
\end{definition}

We use two point to simply our proof co-regular. The first point is to ues the results in \cite{Bouajjani:2015} to ensure FIFO (first in first out) property of each single-priority projections. A execution where only values of one priority is putted is called a single-priority execution. It is obvious that each single-priority projection of sequences in $\seqPQ$ satisfy FIFO property. Let $\textit{toQueue}(e)$ be an execution generated from $e$ by transforming $\textit{put}$ and $\textit{rm}$ into $\textit{enq}$ and $\textit{deq}$, respectively, and then discarding priorities. In Appendix \ref{sec:appendix proof and definition in section definition of co-regular}, according to the result of queue in \cite{Bouajjani:2015}, we construct a set $\textit{Auts}_{\textit{sinPri}}$ of witness automata, and shows that they are enough to ensure that for each data-differentiated executions, each of its single-priority projection without $\textit{rm}(\textit{empty})$ to have ``FIFO'' property, as shown by the following lemma.

@@

According to Lemma \ref{lemma:automata for extended priority queue with single priority}, from now on, it is safe to assume that, for each data-differentiated execution, any of its single-priority projection without $\textit{rm}(\textit{empty})$ has ``FIFO'' property. This effectively exclude some obvious violations, such as $\textit{rm}(a)$ happens before $\textit{put}(a,\_)$.

The second point is to ignore the incomparable priorities. Given a data-differentiated execution $e$, we say that $e$ is a $p$-execution, if $p$ is larger than or equal to the priority of values of $e$. Given a data-differentiated execution $e$ and assume that $p$ is one of the maximal priority of $e$, let $\textit{p-Exec}(e)$ be an execution obtained from $e$ by erasing all operations of values whose priority are incomparable with $\textit{pri}$. The following lemma states that when checking co-regular of $\seqPQ$, it is safe to consider only $p$-executions.

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Given a data-differentiated execution $e$ and $\Gamma\in \{\mathsf{UnmatchedMaxPriority}$ $,\mathsf{MatchedMaxPriority}\}$. $\Gamma\mathsf{\text{-}Conc}(e,x)$ holds, if and only if $\Gamma\mathsf{\text{-}Conc}(\textit{p-Exec}(e),x)$ holds, where $p$ is the priority of $x$.
\end{restatable}
\begin {proof} (Sketch)

To prove the $\textit{only if}$ direction, given $e \sqsubseteq l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \Gamma\mathsf{\text{-}Seq}(s,\alpha)$. Since there is no restriction to $p$-incomparable operations in $u$, $v$ and $w$, we can safely erase all $p$-incomparable values.

To prove the $\textit{if}$ direction, we use a lemma in Appendix, which said that if the projection an execution into a set $O_1$ of operations has linearization $l_1$ and the projection of the execution into $O_2$, the remanning operation, has linearization $l_2$, then the execution has a linearization whose projection into $O_1$ is $l_1$ and projection into $O_2$ is $l_2$.

Given $\textit{p-Exec}(e) \sqsubseteq l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \Gamma\mathsf{\text{-}Seq}(s,\alpha)$. By above lemma we know that $e$ has a linearization $l'$, such that the projection of $l'$ into $p$-incomparable values. It is not hard to see that the priorities comparable with $p$ are either smaller or incomparable with the priorities incomparable with $p$. Thus, $l' \in \seqPQ$. \qed
\end {proof}

Lemma \ref{lemma:pri execution is enough} make our co-regular proof more clear. One consequence of it is that for $e=\textit{p-Exec}(e)$, at most one $\Gamma\mathsf{\text{-}Condition}(e')$ holds.
